A Simple Minibatch Acceptance Test for MCMC

### 2016-05-16 nips_2016.tex
Slight updates to Section 1 and abstract following John's suggestions, also removed my discussion and haoyus discussion on the whole X = g inverse

\begin{document}

\begin{comment}
\end{comment}

\maketitle

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to
perform Metropolis-Hastings tests on an entire dataset. Instead we
describe a novel acceptance test which is applied to minibatches of
data. Previous work on minibatch Metropolis-Hastings use adaptive
batch sizes, which may use far more than one minibatch of data and/or
yield few samples.  Here we describe a simpler test on fixed-sized
minibatches. Instead of adaptive batch size, this test requires a
precondition on the variance of the log proposal probability
ratio. Given a measurement of this variance, minibatch size can be
chosen to support the test in one step, and yielding a number of
samples proportional to the batch/minibatch ratio. Alternatively the
variance condition can be used to adjust the temperature of the
sampling distribution to a ``natural'' value to provide regular
samples at a given minibatch size. The resulting proposal/test is only
slightly more complex than a simple SGD update. In this paper we
derive the test, discuss its implementation, and present several
experiments.

\end{abstract}

@@

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.may
Minibatch updates produce many improvements in the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. MCMC methods in contrast, typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations
- GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed etc.

Here we develop an alternative minibatch acceptance test for MCMC
methods which uses only one fixed-size minibatch per test. Instead of
dynamically adjusting the batch size, we add a prequisite for the test
based on the variance of the log acceptance probability. If this
variance is small enough, the test requires only a single step. This
variance changes slowly on typical problems, and we can estimate it
before performaning the accpeptance test. The minibatch variance
descreases with the minibatch size, so the variance measurement can be
used to set the minimum minibatch size.

There is an additional advantage when using the test for Bayesian
posterior inference. As dataset size grows, the posterior distribution
sharpens and will often concentrate to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become
stuck in one of them (see section 5.1).  The samples are not exploring
the posterior parameter space, and are unlikely to concentrate at a
strong local optimum. They therefore fail to accurately represent a
sample of the posterior because of inadequate search (inadequate
mixing). We argue that it is much more natural to sample the
distribution at a different (higher) temperature determined by the
minibatch size rather than the dataset size. This temperature is
determined in a natural way by the variance prerequisite for our
test. Starting at a high temperature (small minibatch size) which
mixes easily, one can reduce the temperature (increasing minibatch
size) until one reaches the desired target distribution temperature.
Although this approach also involves changing minibath size, the
changes increase minibatch size monotonically, and only a few changes
are performed in the course of a sampling session.
Annealing can also be used for MAP and ML estimation because changing temperature preserves
posterior modes.

The contributions of this paper are as follows:

\begin{enumerate}[noitemsep]
    \item We develop a new minibatch acceptance test which satisfies detailed balance.
    \item We compare performance of our new test and the adaptive sampling method.
    \item Experiments using the approach for posterior estimation.
\end{enumerate}

%{\color{blue}
%Daniel: This section needs work in describing our argument carefully. I will decrease the focus on
%temperature and increase focus on the new MH test once the results are clearer.
%}

@@

Ultimately, minibatch SGHMC and Langevin Dynamics is an orthogonal contribution to ours, two ways of
attempting to solve the same problem of MCMC methods with big datasets. We could augment their
methods with our MH test, to get the benefit of their high quality proposals. Note that while their
methods in theory do not need an MH test, in practice, they need one due to discretization of the
system. Furthermore, tests may also be needed for when the algorithms are run with a constant step
size.

@@

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently: $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \prod_{i=1}^n p(x_i \mid \theta') q(\theta_t \mid
\theta')}{p(\theta_t) \prod_{i=1}^n p(x_i \mid \theta_t) q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log space. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\beta \Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At
any iteration with a current parameter $\theta_t$ and a proposed sample $\theta'$, we can compute
$\Delta$. Let $u$ be a value sampled from uniform distribution $U \sim {\rm Unif}[0,1]$. We accept the candidate if $g(\Delta) > u$, and reject otherwise. This testing process is equivalent to sample a value $x$ from distribution whose CDF is $g(x)$, and accept if $\Delta + x > 0$, reject otherwise. To see why this works,
define a new random variable $X = g^{-1}(U)$ where again, $g$ is the logistic function. Notice that
$X$ has cumulative distribution function $F_X(x) = g(x)$. In other words, the CDF is precisely the
logistic function. This is because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of the uniform random variable here is simply one. Thus, the criteria to accept the
candidate $\theta'$ is equivalent to whether $\Delta > X$. Since $X$ is symmetric about zero, the
acceptance criteria is equivalently expressed as $\Delta + X>0$, as shown in Figure~\ref{fig:part1}.

{\color{blue}
Daniel: The above makes sense, but I \emph{still} don't understand why we had to show that the CDF
of $X$ is the logistic function, because I never use that when I claim that our acceptance test is
now $\Delta > X$. All I need to say is this: we can pre-multiply $g^{-1}$ to $g(\Delta) \ge u$, and
our test is $\Delta > X$. Right???
}
{\color{red}
Haoyu: If you don't show the CDF of $X$ is $g()$, how do you define the $X$. In order to use rule: $\Delta +X >0$, you need to tell people how to get $X$, thus, we need the distribution of $X$. If you just say $Delta > g^{-1}(u)$, then what's X? Then how can you decomposite the $X$ to get the $X_{corr}$? So we need to at least annouce the distribution of $X$ is $g()$.
}

@@

In this section, we use three experiments to reveal the benefits of our mini-batch MH testing. For the first experiment, we illustrate that our mini-batch MH testing can obtain the posterior distribution which is very close to the ideal one. For the second experiment, we use Logistic regression to reveal that our mini-batch sampler is much more efficient than adaptive minibatch approach. The third experiment shows that with the help of temperature, our mini-batch MH testing can be used to obtain the parameters which are close to MAP. 
\subsection{Mixture of Gaussians}\label{ssec:gaussians}

We start with a simple mixture of Gaussians, which has the tied means:

\[
\theta_1 \sim N(0, \sigma_1^2); \theta_2 \sim N(0, \sigma_2^2); x_i \sim \frac{1}{2}N(\theta_1, \sigma_x^2) + \frac{1}{2} N(\theta_2, \sigma_x^2)
\]
where $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. 200 data points are generated from the model with $\theta_1 = 0$ and $\theta_2 = 1$. The theoretical posterior distribution can be seen in Fig.~3a. We use random walk as the proposer. We set the step size 0.1 as a constant. We compare three scenarios: with our mini-batch MH testing (see Fig.~3b), without our mini-batch MH testing (see Fig.~3c), and with the standard MH testing (see Fig.~3d).

@@

From the figures above, we find that with the help of our mini-batch MH testing we can obtain the sampled distribution which are very close to target one. We feature draw our estimation of standard derivation of $X_{\rm norm}$ over iterations, see Fig.~3e. We find with even constant step size, the standard derivation decreases, because the gradient of posterior distribution declines and make the scale of $\Delta'$ deminished when the sampled point is apporching MAP. Thus, the error our transition kernel will declines with even constant step size as discribed in Sect.~ 4.
\subsection{Logistic Regression}\label{ssec:logistic}
In this experiment, we use logistic regression to classify the MNIST dataset. For similification, we only train the binary classifer for digits 1 vs 7. The dataset consisted of 12007 training data and 1000 testing data. Similarly, we use random walk as the proposer. We compare our mini-batch MH testing with the adaptive mini-batch MH testing. Figure 4 shows the loglikelihood value and perdiction accuracy over training process. Since the adaptive MH testing method needs to increase the mini-batch size adaptively, we use the cumulative number of data instance each algorithm consumes as the horizontal axis. We find our mini-batch MH testing can achieve the similar (or even better loglikelihood/prediction accuracy) with consuming much fewer data points. Thus, we can assert that our mini-batch MH testing is much more efficient than the adaptive one.

>>>>

\begin{document}
\maketitle

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to
perform Metropolis-Hastings tests on the full data. To do this, we
describe a novel acceptance test which is applied to minibatches of
data. Previous work on minibatch Metropolis-Hastings use adaptive
batch sizes, which may use far more than one minibatch of data and therefore
yield few samples.  Here we describe a simpler test on fixed-sized
minibatches. Instead of adaptive batch sizes, our test requires a
precondition on the variance of the log proposal probability
ratio. Given a measurement of this variance, a minibatch size can be
chosen to support the test in one step, which yields a number of
samples proportional to the batch/minibatch ratio. Alternatively, the
variance condition can be used to adjust the temperature of the
sampling distribution to a ``natural'' value to provide regular
samples at a given minibatch size. The resulting proposal/test is only
slightly more complex than a simple SGD update. In this paper we
derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, MCMC methods typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations: 
GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed, etc.

In this paper, we develop an alternative minibatch acceptance test for MCMC
methods which uses only one fixed-size minibatch per test. Instead of
dynamically adjusting the batch size, we add a prerequisite for the test
based on the variance of the log acceptance probability. If this
variance is small enough, the test requires only a single step. This
variance changes slowly on typical problems, and we can estimate it
before performing the test. The minibatch variance
decreases as the minibatch size grows, so the variance measurement can be
used to set the minimum minibatch size.

There is another advantage when using our test for Bayesian
posterior inference. As the dataset size grows, the posterior distribution
sharpens and often concentrates to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become
stuck in one of them (see Section~\ref{ssec:gaussians}).  The samples are not exploring
the posterior parameter space, and are unlikely to concentrate at a
strong local optimum. They therefore fail to accurately represent a
sample of the posterior because of inadequate search (i.e.,
mixing). We argue that it is much more natural to sample the
distribution at a higher temperature determined by the
minibatch size rather than the dataset size. This temperature is
determined in a natural way by the variance prerequisite for our
test. Starting at a high temperature (small minibatch size) which
mixes easily, one can reduce the temperature (increasing minibatch
size) until one reaches the desired target distribution temperature.
Although this approach also involves changing minibatch size, the
changes increase minibatch size monotonically, and only a few changes
are performed in the course of a sampling session.
Annealing can also be used for MAP and ML estimation because changing temperature preserves
posterior modes.

To be precise, the contributions of this paper are as follows:

\begin{enumerate}[noitemsep]
    \item We develop a new minibatch acceptance test which satisfies detailed balance.
    \item We compare performance of our new test and the adaptive sampling method.
    \item We experiment using the approach for posterior estimation.
\end{enumerate}

@@

Minibatch SGHMC and Langevin Dynamics are orthogonal to our work; two ways of attempting
to solve the same problem of MCMC methods with big datasets. We could augment their methods with our
MH test, to get the benefit of their high quality proposals. Note that while their methods in theory
do not need an MH test, in practice, they need one due to discretization of the system. Furthermore,
    tests may also be needed for when the algorithms are run with a constant step size.

@@

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \prod_{i=1}^n p(x_i \mid \theta') q(\theta_t \mid
\theta')}{p(\theta_t) \prod_{i=1}^n p(x_i \mid \theta_t) q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log space. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\beta \Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the
condition in Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily
sample from it using the following procedure. At any iteration with a current parameter $\theta_t$
and a candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u
\sim {\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. To see why, define $X = g^{-1}(u)$ where
again, $g$ is the logistic function. The CDF of $X$ is precisely the logistic function because for
arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of the uniform random variable here is simply one. Thus, the criteria to accept the
candidate $\theta'$ is equivalent to whether $\Delta > X$. Since $X$ is symmetric about zero, the
acceptance criteria is equivalently expressed as $\Delta + X>0$, as shown in Figure~\ref{fig:part1}.

@@

In this section, we use three experiments to reveal the benefits of our mini-batch MH testing. For
the first experiment, we illustrate that our mini-batch MH testing can obtain the posterior
distribution which is very close to the ideal one. For the second experiment, we use Logistic
regression to reveal that our mini-batch sampler is much more efficient than adaptive minibatch
approach. The third experiment shows that with the help of temperature, our mini-batch MH testing
can be used to obtain the parameters which are close to MAP.

\subsection{Mixture of Gaussians}\label{ssec:gaussians}

{\color{blue}
Daniel: TODO Have to show that with more data, the posterior peaks (because it is referenced by the
introduction).
}

We start with a simple mixture of Gaussians, which have tied means:
\[
\theta_1 \sim N(0, \sigma_1^2); \theta_2 \sim N(0, \sigma_2^2); x_i \sim \frac{1}{2}N(\theta_1, \sigma_x^2) + \frac{1}{2} N(\theta_2, \sigma_x^2)
\]
where $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. 200 data points are generated from the
model with $\theta_1 = 0$ and $\theta_2 = 1$. The theoretical posterior distribution can be seen in
Fig.~3a. We use random walk as the proposer. We set the step size 0.1 as a constant. We compare
three scenarios: with our mini-batch MH testing (see Fig.~3b), without our mini-batch MH testing
(see Fig.~3c), and with the standard MH testing (see Fig.~3d).

@@

From the figures above, we find that with the help of our mini-batch MH testing we can obtain the
sampled distribution which are very close to target one. We feature draw our estimation of standard
derivation of $X_{\rm norm}$ over iterations, see Fig.~3e. We find with even constant step size, the
standard derivation decreases, because the gradient of posterior distribution declines and make the
scale of $\Delta'$ deminished when the sampled point is apporching MAP. Thus, the error our
transition kernel will declines with even constant step size as discribed in Sect.~ 4.

\subsection{Logistic Regression}\label{ssec:logistic}

In this experiment, we use logistic regression to classify the MNIST dataset. For similification, we
only train the binary classifer for digits 1 vs 7. The dataset consisted of 12007 training data and
1000 testing data. Similarly, we use random walk as the proposer. We compare our mini-batch MH
testing with the adaptive mini-batch MH testing. Figure 4 shows the loglikelihood value and
perdiction accuracy over training process. Since the adaptive MH testing method needs to increase
the mini-batch size adaptively, we use the cumulative number of data instance each algorithm
consumes as the horizontal axis. We find our mini-batch MH testing can achieve the similar (or even
better loglikelihood/prediction accuracy) with consuming much fewer data points. Thus, we can assert
that our mini-batch MH testing is much more efficient than the adaptive one.
